The Design and Implementation of the Synchronous Language \CEU

### 2016-10-04 paper.tex
(>) many fixes

\begin{abstract}
\CEU is a synchronous language inspired by Esterel with a simpler semantics and 
more fine-grained control over program execution targeting soft real-time
systems.
%
\CEU uses an event-triggered notion of time that enables compile-time checks to 
detect conflicting concurrent statements, resulting in deterministic and 
concurrency-safe programs.
%
Using Esterel as a base, we present the particularities of our design, such as 
stack-based internal events, concurrency detection, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics for soft real-time systems with fine-grained control
for intra-reaction execution, which is amenable to concurrency checks that
improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
In contrast with Esterel's semantics, this prevents programs to be compiled
directly and efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
for hard real-time systems~\cite{esterel.fixed}.

@@

Figure~\ref{lst.syn} presents the subset of the concrete syntax of \CEU used in
the paper as a quick reference.

@@

input  <type> <ids>;                        // external input  events
output <type> <ids>;                        // external output events
event  <type> <ids>;                        // internal events
var    <type> <id> = <exp>;                 // a variable with initial value

@@

<id> = await <id>;                          // await an event and assign received value
<id> = await <time>;                        // await time and assign delayed delta
emit <id> => <exp>;                         // emit an event passing a value

@@

<stmt> ; <stmt>                             // sequence
if <exp> then <stmts> else <stmts> end      // conditional
loop do <stmts> end                         // repetition
do <stmts> end                              // lexical block

@@

<_id>(<exps>)                               // C call (identifier starts with `_')
native do <stmts> end                       // declarations in C
native @const <_ids>                        // annotations: constant symbols
native @pure  <_ids>                        //              pure functions
native @safe  <_ids> with <_ids>            //              non-conflicting symbols
\end{lstlisting}
\caption{ Subset of the concrete syntax of \CEU used in the paper.
\label{lst.syn}
}
\end{figure}

In the subsections that follow, we discuss the main differences between \CEU and
Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static concurrency checks
    (\ref{sec.ceu.checks});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish the section with a summary of our design
    (\ref{sec.ceu.summary}).
%
We present the formal specification of the semantics of \CEU with a simplified
abstract syntax in a separate paper~\cite{ceu.phd}.

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} assumes an external observer with an absolute
reference clock that timestamps event occurrences over a
continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
$17ms521us$).
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU reacts to the two instances of \code{E} independently, which are
handled in sequence.
%
%Note that, by definition, each reaction corresponds to an event occurrence of 
%the same length.
\end{itemize}

Sample-driven execution maps directly to hardware implementations and
involve no runtime queues.
%
However, considering the soft real-time application domain of \CEU, we
decided for the unique and queue-based semantics in \CEU for the reasons
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is implementation dependent:}
Tick lengths are not part of the program specification.
For instance, if we log the \code{Real World} box timeline and reproduce it in
implementations with different tick lengths, the behaviors might diverge.
%
\item \textbf{Events are never absolutely simultaneous:}
We consider that the notion of simultaneity should not be imposed by the
language, but defined explicitly for each use case (to be discussed in
Section~\ref{sec.ceu.timers}).
%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
%first-class timers, we show an example to illustrate how to detect simultaneous 
%button clicks in \CEU.
In tick-based approaches, simultaneity depends on the length of discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
events \code{B} and \code{C} are simultaneous, even though \code{A} and
\code{B} ``physically'' happen much closer to one another.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to exclusive events are atomic and never overlap.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

Note that Esterel supports declarations for mutually exclusive inputs which
cannot be simultaneous in the same reaction
(e.g., ``\code{relation A\#B\#C\#D\#E;}'')~\cite{esterel.primer}.
Making all inputs mutually exclusive and adopting the variable-length-tick
approach is equivalent to \CEU's notion of time, which can be seen as an
imposed restriction on the semantics of Esterel.

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a tradeoff that an \code{await} is only
valid for the next reaction, i.e., if an \code{await} and \code{emit} occur 
simultaneously in parallel trails, the \code{await} does not awake.
%
This approach avoids any form of graph construction or data dependency analysis
as required in Esterel, e.g.:
\emph{``set or reset of a signal must always precede any test of this signal
(signal coherence law and constructivity''}~\cite{esterel.saxort}.
%
Also, these \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution by design.
Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
However, if the \code{emit e} (ln. 6) could awake the \code{await e} 
instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
\code{loop} instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by placing the code to execute before the \code{await}.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

\begin{figure}[b]
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    ##define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{\ax Definitions and uses of symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id();
native @safe  _f() with _g();

@@

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels, which makes the calls non commutative and prevents the code 
to compile.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
\code{\_redraw\_non\_commutative} (ln. 2) and annotate it as safe (ln. 4--5) to
make the code compile successfully.
The new function name is explicit about its effect, improving code readability.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
(ln. \ax:7).
\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize} construct.

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
Due to platform overheads and occasional slow reactions, an \code{await 10ms}
suspends a line of execution for \emph{at least} 10 milliseconds.
In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy and takes 15ms to notify 
\CEU.
The scheduler will notice that the \code{await 10ms} (ln. 2) has not only 
already expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1} (ln. 3), and invokes 
\code{await 1ms} (ln. 4).
As the current delta is still higher than the requested timeout (i.e. $5ms > 
1ms$), the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

The compile-time concurrency checks phase detects inconsistencies in \CEU 
programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.checks}.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
As far as we know, \CEU is the first to encrust this event-triggered notion of 
time in the core of the language, which is a prerequisite for the concurrency 
checks that enable safe shared-memory concurrency.

>>>>

\begin{abstract}
\CEU is a synchronous language targeting soft real-time systems.
It is inspired by Esterel and has a simple semantics with fine-grained control
over program execution.
%
\CEU uses an event-triggered notion of time that enables compile-time checks to 
detect conflicting concurrent statements, resulting in deterministic and 
concurrency-safe programs.
%
We present the particularities of our design in comparison to Esterel, such as 
stack-based internal events, concurrency checks, safe integration with \C, 
and first-class timers.
%
We also present two implementation back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
\end{abstract}

@@

In this work, we present \CEU, a new programming language that inherits the 
synchronous and imperative mindset of Esterel but diverges in some fundamental 
semantic aspects.
%
\CEU proposes a semantics for soft real-time systems with fine-grained control
for intra-reaction execution which is amenable to concurrency checks that
improve safety.
%
The list that follows summarizes the contributions behind the design of \CEU:
%
\begin{itemize}
%
\item Unique and queue-based external events, which define the notion of time 
in \CEU.
% (Section~\ref{sec.ceu.exts}).
%
\item Stack-based internal events for intra-reaction communication, which also
provides a limited form of coroutines.
% (Section~\ref{sec.ceu.ints}).
%
\item Static concurrency checks to detect suspicious concurrent statements.
%(Section~\ref{sec.ceu.checks}).
%
\item Safe integration with \C that enforces finalization for external 
resources.
% (Section~\ref{sec.ceu.c}).
%
\item First-class timers with dedicated syntax and automatic synchronization.
%(Section~\ref{sec.ceu.timers}).
\end{itemize}
%
We also present a lightweight single-threaded implementation of \CEU with two 
back ends:
one aiming for resource efficiency and interoperability with \C,
and another as a virtual machine that allows remote reprogramming.
%
Our implementations target resource-constrained devices, such as \emph{Arduino} 
and \emph{MICAz} sensor nodes based on 8-bit microcontrollers%
\footnote{
Both \emph{Arduino} and \emph{MICAz} use the 8-bit \emph{ATmega328} 
microcontroller with 32K of FLASH and 2K of SRAM.
}, showing a practical aspect of the proposed semantics.
%:\url{http://www.atmel.com/devices/atmega328.aspx}}

@@

The semantics of \CEU implies an implementation with a runtime stack and a
sequential scheduler.
In contrast with Esterel's semantics, this prevents programs to be compiled
directly and efficiently to hardware~\cite{esterel.constructive}.
It also makes difficult to statically determine worst-case reaction times
towards hard real-time systems~\cite{esterel.fixed}.

@@

Figure~\ref{lst.syn} shows the subset of the concrete syntax of \CEU used in
this paper as a quick reference.
%
In a separate document, we present the formal semantics with a simplified
abstract syntax~\cite{ceu.phd}.

@@

input  <type> <ids>;                        // external input  events
output <type> <ids>;                        // external output events
event  <type> <ids>;                        // internal events
var    <type> <id> = <exp>;                 // a variable with an initial value

@@

<id> = await <id>;                          // await an event and assign the received value
<id> = await <time>;                        // await time and assign the delayed delta
emit <id> => <exp>;                         // emit an event passing a value

@@

<stmt> ; <stmt>                             // sequence
if <exp> then <stmts> else <stmts> end      // conditional
loop do <stmts> end                         // repetition
do <stmts> end                              // explicit block

@@

<_id>(<exps>)                               // C call (identifier starts with `_')
native do <stmts> end                       // declarations in C
native @const <_ids>                        // annotations: constant symbols
native @pure  <_ids>                        //              pure functions
native @safe  <_ids> with <_ids>            //              non-conflicting symbols
\end{lstlisting}
\caption{ Subset of the concrete syntax of \CEU used in this paper.
\label{lst.syn}
}
\end{figure}

In the subsections that follow, we discuss the main differences between \CEU and
Esterel:
%
Unique and queue-based external events
    (\ref{sec.ceu.exts});
Stack-based internal events
    (\ref{sec.ceu.ints});
Static concurrency checks
    (\ref{sec.ceu.checks});
Safe integration with \C
    (\ref{sec.ceu.c});
and
First-class synchronized timers
    (\ref{sec.ceu.timers}).
We finish the section with a summary of our design
    (\ref{sec.ceu.summary}).

@@

Figure~\ref{fig.ticks} compares the discrete notions of time in two variations 
of Esterel and in \CEU.
The box \code{Real World} assumes an external observer with an absolute
reference clock which timestamps event occurrences over a
continuous timeline~\cite[Chapter~3]{es.time} (e.g., event \code{C} occurs at
$17ms521us$).
The other boxes show how the same occurring events fit differently in each 
logical notion of time.
%
\begin{itemize}
\item \code{[Box-1]}:
\emph{Esterel with fixed-length ticks}~\cite{esterel.fixed}.
%
(Usually referred as \emph{sample-driven execution}~\cite{rp.twelve}.)
%
We assume a reaction \code{R(boot)} at \code{tick-0} which happens before any 
input.
%
The input \code{A} ``physically'' occurs during the boot reaction but, because 
time is discrete, its corresponding reaction only executes in the next tick.
%
Note that \code{R(A)} takes more time than \code{tick-1} and invades 
\code{tick-2}, causing a \emph{timing violation}~\cite{esterel.fixed}.
%
The events \code{B} and \code{C} occur during \code{tick-1} and are delayed to 
happen \emph{simultaneously} at \code{tick-2} with \code{R(B+C)}.
%
Since no new events occur during \code{tick-2}, the CPU stays idle during the 
whole \code{tick-3}.
%
Finally, one instance of event \code{D} and two instances of event \code{E} 
occur during the idle \code{tick-3}.
However, only one occurrence of \code{E} can be considered in \code{R(D+E)}.
%
\item \code{[Box-2]}:
\emph{Esterel with variable-length ticks}~\cite{esterel.variable}.
This approach avoids the timing violation for \code{R(A)} and also results in
smaller idle periods because it adjusts the tick lengths to match the CPU times 
for the reactions.
For instance, the occurrence of \code{D} interrupts the idle \code{tick-3} to 
react alone as \code{R(D)} at \code{tick-4}.
Similarly to the fixed-tick approach, only one of the two simultaneous 
occurrences of \code{E} is considered at \code{tick-5}, now because \code{R(D)} 
takes too long.
%
\item \code{[Box-3]}:
\emph{\CEU with unique and queue-based input events}.
We also assume a reaction \code{R(boot)} before any input.
%
Because the occurrence of event \code{A} is unique during \code{R(boot)}, the 
behavior in \CEU is similar to \code{Box-2} for its first two reactions 
(\code{tick-0} and \code{tick-1}).
%
However, \CEU does not consider the events \code{B} and \code{C} as 
simultaneous, and handles each in subsequent reactions \code{R(B)} and 
\code{R(C)}.
%
We assume the CPU times for \code{R(B+C)} in Esterel and \code{R(B)+R(C)} in 
\CEU to be roughly the same.
This way, the first idle periods in \code{Box-2} and \code{Box-3} coincide.
%
Finally, \CEU reacts to the two instances of \code{E} independently, which are
handled in sequence.
%
%Note that, by definition, each reaction corresponds to an event occurrence of 
%the same length.
\end{itemize}

Sample-driven execution maps directly to hardware implementations and
involve no runtime queues.
%
However, considering the soft real-time application domain of \CEU, we
decided for the unique and queue-based semantics in \CEU for the reasons
that follow:
%
\begin{itemize}
\item \textbf{A ``tick'' is implementation dependent:}
Tick lengths are not part of the program specification.
For instance, if we log the \code{Real World} box timeline and reproduce it in
implementations with different tick lengths, the behaviors might diverge.
%
\item \textbf{Events are never absolutely simultaneous:}
We consider that the notion of simultaneity should be defined explicitly for
each use case (to be discussed in Section~\ref{sec.ceu.timers}).
%In Section~\ref{sec.ceu.timers}, after introducing internal events and 
%first-class timers, we show an example to illustrate how to detect simultaneous 
%button clicks in \CEU.
In tick-based approaches, simultaneity depends on the length of discrete ticks.
For instance, in \code{box-1} and \code{box-2} of Figure~\ref{fig.ticks}, 
events \code{B} and \code{C} are simultaneous, even though \code{A} and
\code{B} ``physically'' happen much closer to one another.
%
\item \textbf{Unique input events imply mutual exclusion:}
Reactions to exclusive events are atomic and never overlap.
Automatic mutual exclusion simplifies reasoning about concurrency and is a 
prerequisite for the concurrency checks to be discussed in 
Section~\ref{sec.ceu.checks}.
\end{itemize}

Note that Esterel supports declarations for mutually exclusive inputs that
cannot be simultaneous in the same reaction
(e.g., ``\code{relation A\#B\#C\#D\#E;}'')~\cite{esterel.primer}.
Making all inputs mutually exclusive and adopting the variable-length-tick
approach is equivalent to \CEU's notion of time, which can be seen as an
imposed restriction on the semantics of Esterel.

@@

Another distinction regarding event handling in comparison to \CEU is that 
Esterel supports same-cycle bi-directional 
communication~\cite{esterel.compiling}, i.e., two threads can react to one 
another during the same cycle due to mutual signal dependencies
(a.k.a. \emph{instantaneous dialogue}~\cite{rp.book}).
%
\CEU takes a different approach, posing a restriction that an \code{await} is
only valid for the next reaction, i.e., if an \code{await} and \code{emit}
occur simultaneously in parallel trails, the \code{await} does not awake.
%
These \emph{delayed awaits} avoid corner cases of instantaneous
termination and re-execution of statements in the same reaction (known as
\emph{schizophrenic statements}~\cite{esterel.constructive}).
%
Also, this approach does not require data dependency analysis like Esterel
(e.g., \emph{``(set or reset of a signal must always precede any test of this
signal (signal coherence law and constructivity)''}~\cite{esterel.saxort}).

@@

The example in Figure~\ref{lst.await} illustrates delayed awaits, which 
prevents infinite execution in the loop.
Both sides of the \code{par/or} have an \code{await} statement (ln. 4,7), which 
characterizes the enclosing \code{loop} as non instantaneous (ln 2--9).
However, if the \code{emit e} (ln. 6) could awake the \code{await e} 
instantaneously (ln. 4), the \code{par/or} would terminate and restart the 
\code{loop} also instantaneously, resulting in infinite execution.
%
\begin{comment}
\begin{figure}
%\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
event void e;
par do
    loop do
        <...>     // code that awaits some period
        emit e;   // periodic request
    end
with
    loop do
        <...>     // code to execute immediately and then periodically
        await e;  // await after
    end
end
\end{lstlisting}
%\end{minipage}
%
%\begin{minipage}[t]{0.50\linewidth}
%\begin{lstlisting}[numbers=left,xleftmargin=3em]
%event void e;
%var bool should_execute = false;
%par do
%    loop do
%        <...>         // code that awaits
%        if <...> then // some conditon
%            should_execute = true;
%            emit e;
%        end
%    end
%with
%    loop do
%        if should_execute then
%            <...>     // code to execute
%        end
%        await e;
%    end
%end
%\end{lstlisting}
%\end{minipage}
%
\caption{ An example that circumvents the \emph{delayed await} by post-fixing 
the \code{await} inside the \code{loop}.
% (in the left), and by copying the condition test (in the right).
\label{lst.delay}
}
\end{figure}
\end{comment}
%
In atypical scenarios requiring immediate awake, delayed awaits can be 
circumvented by placing the code to execute before the \code{await}.
%
\begin{comment}
For instance, sometimes we need to execute a block of code immediately, and 
then periodically from internal event requests, as illustrated in 
Figure~\ref{lst.delay}.
In this case, the \code{await} moved to the end of the loop (ln. 10) makes the 
periodic code to also execute immediately (ln. 9), and then in reactions to 
each \code{emit} request (ln. 5).
If the periodic \code{emit} depends on a condition, then the code 
transformation becomes more intricate, requiring an extra condition test around 
the periodic code to prevent its immediate execution.
\end{comment}
%
On the one hand, we transfer the burden of dealing with these corner cases to 
the programmer.
On the other hand, we simplify the semantics of the language and eliminate the 
need for complex analysis to deal with schizophrenic statements.

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A, B;
var int x = 1;
par/and do
    await A;
    x = x + 1;
with
    await B;
    x = x * 2;
end
\end{lstlisting}
\centering\small{\ax Accesses to \code{x} are safe}
\end{minipage}
%
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
input void A;
var int y = 1;
par/and do
    await A;
    y = y + 1;
with
    await A;
    y = y * 2;
end
\end{lstlisting}
\centering\small{\bx Accesses to \code{y} are unsafe}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure}

@@

\begin{figure}
\begin{minipage}[t]{0.50\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3.5em]
native do
    #define NUM 10
    void f  (void)  { <...> }
    void g  (int v) { <...> }
    int  id (int v) { <...> }
end
par/and do
    _f();
with
    _g(_id(_NUM));
end
\end{lstlisting}
\centering\small{\ax Definitions and uses of symbols}
\end{minipage}
%
\begin{minipage}[t]{0.44\linewidth}
\begin{lstlisting}
native @const _NUM;
native @pure  _id();
native @safe  _f() with _g();

@@

From our experience, however, we find that programs often need non-commutative 
concurrent calls.
%
This is the case for logging (e.g., calls to \code{\_printf} in trails in 
parallel) and for redrawing objects in the screen.
%
Figure~\ref{lst.redraw}.a shows an abstract code to animate and redraw the 
objects \code{background} and \code{foreground} in trails in parallel.
In typical graphical APIs, consecutive calls to \code{\_redraw} overwrites 
conflicting pixels, which makes the calls non commutative and prevents the code 
to compile.
%
However, in this case we want to rely on lexical order to always redraw the 
\code{background} object before the \code{foreground} object.
%
Therefore, in Figure~\ref{lst.redraw}.b, we redefine \code{\_redraw} as
\code{\_redraw\_non\_commutative} (ln. 2), communicating its effect explicitly,
and annotate it as safe (ln. 4--5) to make the code compile successfully.

@@

\CEU also enforces the use of \code{finalize} for system calls that deal with
pointers representing resources, as illustrated in the two examples of
Figure~\ref{lst.fin.ceu}:
%
\begin{itemize}
\item If \CEU \textbf{passes} a pointer to a system call (ln. \ax:5), the
pointer represents a \textbf{local} resource (ln. \ax:2) that requires finalization
(ln. \ax:7).
\item If \CEU \textbf{receives} a pointer from a system call return (ln. \bx:4),
the pointer represents an \textbf{external} resource (ln. \bx:2) that requires
finalization (ln. \bx:6).
\end{itemize}
%
\CEU tracks the interaction of system calls with pointers and requires 
finalization clauses to accompany them.
%
In the example in Figure~\ref{lst.fin.ceu}.a, the local variable \code{msg} 
(ln. 2) is an internal resource passed as a pointer to \code{\_send\_request} 
(ln. 5), which is an asynchronous call that transmits the buffer in the 
background.
If the block aborts (ln. 11) before receiving an acknowledge from the 
environment (ln. 9), the local \code{msg} goes out of scope and the external 
transmission now holds a \emph{dangling pointer}.
The finalization ensures that the transmission also aborts (ln. 7).
%
In the example in Figure~\ref{lst.fin.ceu}.b, the call to \code{\_fopen} (ln.  
4) returns an external file resource as a pointer.
If the block aborts (ln. 12) during the \code{await A} (ln. 9), the file 
remains open as a \emph{memory leak}.
The finalization ensures that the file closes properly (ln. 6).
%
In both cases, the code does not compile without the \code{finalize}
construct.%
\footnote{
The compiler only enforces the programmer to write the finalization clause, but
cannot check if it actually handles the resource properly.
}

@@

The \code{await} statement of \CEU supports wall-clock time and handles deltas 
automatically, resulting in more robust applications.
In the example in Figure~\ref{lst.timers}.a, suppose that after the first 
\code{await} request, the underlying system gets busy%
\footnote{Due to platform overheads (e.g., OS scheduling) and occasional slow
reactions, \code{await~10ms} means suspending a line of execution for
\emph{at least} 10 milliseconds.
}
and takes 15ms to notify 
\CEU.
The scheduler will notice that the \code{await~10ms} (ln. 2) has not only 
already expired, but is delayed with \code{delta=5ms}.
Then, the awaiting trail awakes, sets \code{v=1} (ln. 3), and invokes 
\code{await 1ms} (ln. 4).
As the current delta is still higher than the requested timeout (i.e. $5ms > 
1ms$), the trail is rescheduled for execution, now with \code{delta=4ms}.

@@

The compile-time concurrency checks detect inconsistencies in \CEU programs.
Here, we focus on the algorithm that detects suspicious concurrent statements, 
such as accesses to shared variables, as discussed in 
Section~\ref{sec.ceu.checks}.

@@

\CEU has a strong influence from Esterel but differ in the most fundamental 
aspect of the notion of time.
%
In \CEU, instead of clock ticks, atomic external event occurrences that define 
time units.
%
The event-driven approach of \CEU is widespread~\cite{sync_async.whynotthreads} 
and popular in many software communities, such as web frameworks (e.g., 
\emph{jQuery}~\cite{js.jquery} and \emph{Node.js}~\cite{js.node}), GUI toolkits 
(e.g., \emph{Tcl/Tk}~\cite{tcl.tk} and \emph{Java Swing}~\cite{java.swing}), 
and Games~\cite{gamepatterns}.
%
\CEU encrusts this event-triggered notion of time in the core of the language,
which is a prerequisite for the concurrency checks that enable safe
shared-memory concurrency.
